The researcher’s major research question is: How have house prices in U.S. states changed over the last few decades, and are changes in prices related to population in some way?  

DATA
** CPI : Consumer price index
  The consumer price index (CPI) is the instrument to measure inflation. It is used to estimate the average variation between two given periods in the prices of products consumed by households. It is a composite measurement of trends in the prices of products, at constant quality. 

** state-abbrevs : two letter code and census region for each state in the US
** Freddie Mac House Price Index : Tracking the average house price in each state since 1975
** Census Data : population data.


```{r}

library(readxl) 
library(gapminder)
cb_palette = c("#999999", "#E69F00", "#56B4E9", "#009E73",
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
library(ggplot2)

house_price = read_excel("State_and_US_SA.xls", range = "A6:BA546")
cpi = read.csv("cpi.csv", header = TRUE)


#Converting month to date-time
house_price$Month =  apply(house_price, 1, function(row) as.Date(paste(gsub("M","-",row[1]),"-01", sep="")))
house_price$Month = as.Date(house_price$Month, origin='1970-01-01')
house_price$Year = as.integer(format(house_price$Month, "%Y"))


#Converting month to date-time
cpi$Month = paste(cpi$Year, cpi$Period,sep='')
cpi$Month = apply(cpi, 1, function(row) as.Date(paste(gsub("M","-",row[6]),"-01", sep="")))
cpi$Month = as.Date(cpi$Month, origin='1970-01-01')
cpi_new = subset(cpi, select = c(Value, Month))

#Merging the two dataframes
merged_df = merge(cpi_new, house_price, by = 'Month')
colnames(merged_df)[2] = "CPI"

which(merged_df$Month == "2000-12-01") #312

# inflation adjusted house price
P = merged_df[3:53] 
p2 = P/ merged_df$CPI 
p2[312,]
C = (100/t(p2[312,]))
merged_df[3:53] = p2*C 
merged_df[312,] # make the house price of each state in 2000 Dec to 100

merged_df = aggregate(.~ Year, merged_df, FUN=mean)[,-2]
colnames(merged_df)[54] = 'Adjusted'
merged_df

```


1. House prices over time: How have house prices in the U.S changed from 1975 to 2019, after adjusting for inflation, i.e. relative to the CPI? (We stop in 2019 to avoid having to worry about COVID effects.) 
1) How have changes in prices varied by state? 
Which states have seen the biggest increases in real house prices, 
2) and which have seen the biggest decreases? 

3) Have changes in prices within each state mostly followed the same basic pattern, 
and are there outliers to that pattern? 

4) Do the typical patterns vary between the four regions (Northeast, Midwest, South, and West)?
```{r}

# Transporting to long format for analysis
merged_df_long <- merged_df[-1,] |> 
  pivot_longer(!c(Year,  Adjusted, CPI), 
               names_to = "state", values_to = "HousePrice")

#graph showing change of the house price from 1975 to 2019
merged_df_long$Year = as.numeric(merged_df_long$Year)
A = ggplot() + geom_line(data = merged_df_long, 
                         aes(x = Year, y = HousePrice, group = state, labels(state)), alpha = 0.25)+ geom_point() +
  ggtitle("House price for each state over the year") + 
  scale_x_continuous(breaks = c(1975, 1985, 1995, 2005, 2015, 2025 )) +
  geom_text(data=subset(merged_df_long, HousePrice > 250 | HousePrice < 51),
            aes(Year,HousePrice,label=state))
A  

# facet by each state
A + facet_wrap(~state)

summary(merged_df_long$HousePrice)
 
```

follows sign curve, peak around 1978, 1987, 2008, down again around 2010 
housing price index start around 100 at the end of year of 2000 thus they converge

## region
```{r}


region = read.delim("state_abbrevs.txt", sep="", header= TRUE)
colnames(region)[2] = 'state' 

# adding regions (Northeast, Midwest, South, and West)
merged_region = merge(merged_df_long,region[-1], by='state')

# region mean
only_region = merged_region[,-c(3,4)]
house_price_region=  only_region |> group_by(Region, Year) |> summarise(HousePrice = mean(HousePrice)) 

B = ggplot() + geom_line(data = house_price_region, aes(x = Year, y = HousePrice, group = Region, color = Region), alpha = 1)+ ggtitle("House price for each region over the year") + scale_x_continuous(breaks = c(1975, 1985, 1995, 2005, 2015, 2025 )) 
B
```

## states within a region
```{r}
West <- filter(merged_region, Region == "West")
South <- filter(merged_region, Region == "South")
Northeast <- filter(merged_region, Region == "Northeast")
Midwest <- filter(merged_region, Region ==  "Midwest")

W = ggplot(West, aes(y = HousePrice, x = Year, color = state)) +
  geom_line() + ggtitle("House price for West region over the year") + 
  scale_x_continuous(breaks = c(1975, 1985, 1995, 2005, 2015, 2025 )) 

S = ggplot(South, aes(y = HousePrice, x = Year, color = state)) +
  geom_line() + ggtitle("House price for South region over the year") + 
  scale_x_continuous(breaks = c(1975, 1985, 1995, 2005, 2015, 2025 ))

NE = ggplot(Northeast, aes(y = HousePrice, x = Year, color = state)) +
  geom_line() + ggtitle("House price for Northeast region over the year") + 
  scale_x_continuous(breaks = c(1975, 1985, 1995, 2005, 2015, 2025 ))

MW = ggplot(Midwest, aes(y = HousePrice, x = Year, color = state)) +
  geom_line() + ggtitle("House price for Midwest region over the year") + 
  scale_x_continuous(breaks = c(1975, 1985, 1995, 2005, 2015, 2025 ))

W
S
NE
MW

```
