---
title: "EDA_miniProject"
author: "Hyeseon Seo"
date: "2023-09-14"
output: pdf_document
---

The researcher’s major research question is: How have house prices in U.S. states changed over the last few decades, and are changes in prices related to population in some way?  

DATA
** CPI : Consumer price index
  The consumer price index (CPI) is the instrument to measure inflation. It is used to estimate the average variation between two given periods in the prices of products consumed by households. It is a composite measurement of trends in the prices of products, at constant quality. 

** state-abbrevs : two letter code and census region for each state in the US
** Freddie Mac House Price Index : Tracking the average house price in each state since 1975
** Census Data : population data.


```{r}
library(tidyverse)
library(readxl) 
library(gapminder)
cb_palette = c("#999999", "#E69F00", "#56B4E9", "#009E73",
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
library(ggplot2)

house_price = read_excel("State_and_US_SA.xls", range = "A6:BA546")
cpi = read.csv("cpi.csv", header = TRUE)

house_price$Month =  apply(house_price, 1, function(row) as.integer(unlist(strsplit(row["Month"], "M"))[1]))
colnames(house_price)[1]= "Year"
house_price = aggregate(.~ Year, house_price, FUN=mean)

#cpi$Month = paste(cpi$Year, cpi$Period,sep='')
cpi_new = subset(cpi, select = c(Value, Year))
cpi_new = aggregate(Value~ Year, cpi_new, FUN=mean)
merged_df = merge(cpi_new,house_price,by='Year')
colnames(merged_df)[2] = "CPI"

# inflation adjusted house price
#merged_df$CPI_factor = merged_df$CPI / lag(merged_df$CPI)
P = merged_df[3:54] 
p2 = P/ merged_df$CPI 
C = (100/t(p2[26,]))
merged_df[3:54] = p2*C
merged_df[26,] # make the house price of each state in 2000 around 100


colnames(merged_df)[54] = 'Adjusted'

```




1. House prices over time: How have house prices in the U.S changed from 1975 to 2019, after adjusting for inflation, i.e. relative to the CPI? (We stop in 2019 to avoid having to worry about COVID effects.) 
1) How have changes in prices varied by state? 
Which states have seen the biggest increases in real house prices, 
2) and which have seen the biggest decreases? 

3) Have changes in prices within each state mostly followed the same basic pattern, 
and are there outliers to that pattern? 

4) Do the typical patterns vary between the four regions (Northeast, Midwest, South, and West)?
```{r}

# Transporting to long format for analysis
merged_df_long <- merged_df[-1,] |> 
  pivot_longer(!c(Year,  Adjusted, CPI), 
               names_to = "state", values_to = "HousePrice")

#graph showing change of the house price from 1975 to 2019
merged_df_long$Year = as.numeric(merged_df_long$Year)
A = ggplot() + geom_line(data = merged_df_long, 
                         aes(x = Year, y = HousePrice, group = state, labels(state)), alpha = 0.25)+ geom_point() +
  ggtitle("Figure 1. House price for each state over the year") + 
  scale_x_continuous(breaks = c(1975, 1985, 1995, 2005, 2015, 2025 )) +
  geom_text(data=subset(merged_df_long, HousePrice > 360 | HousePrice < 17),
            aes(Year,HousePrice,label=state))
A

# facet by each state
wrap_plot <- A + facet_wrap(~state) +
  scale_x_continuous(breaks = c(1975, 1997, 2019)) +
  scale_y_continuous(breaks = c(50, 250)) +
  labs(
    title = "Figure 2. Housing Price Index for Each State (1975-2019)"
  )
wrap_plot

summary(merged_df_long$HousePrice)
 
```

In Figure 1, a majority of states appear to feature an HPI decrease that begins roughly around 1980. Some states begin to reverse course and feature an increase in the latter-half of the decade. Setting the baseline at 100 HPI is can also be seen from the graph. The most consistent feature from Figure 1 is the housing bubble in the early oughts. This is where housing prices were increasing and investors could sell houses for more money, even if borrowers defaulted. Lenders made sub prime mortgages to low-income borrowers. Borrowers' inability to keep up with payments every months led to foreclosures and more houses were back on the market. However, the demand was down and couldn't meet the supply, so the price of homes were collapsing. This burst is what we see in every state around 2006-2007 and was an important cause to the Great Recession in the U.S. 

We can see how each state behaved in Figure 2. Many states in the Midwest and upper South regions in the United States feature roughly flat curves (e.g., Tennessee and Kentucky). Some states feature a dramatic increase in HPI, such as D.C. and Hawaii. These states are relatively small, so there may be a connection with population. Although, Oregon and Virginia feature HPI increases as well. The states that feature decreases are rural states out West like Kansas and Oklahoma, and rural southern states such as West Virginia and Mississippi. Some states reacted more dramatically to the Bubble and Recession than others. While many states remained flat, states like Nevada, California, and Arizona witnessed steep decreases after the Housing Crisis of 2006. 

## region
```{r}


region = read.delim("state_abbrevs.txt", sep="", header= TRUE)
colnames(region)[2] = 'state' 
region

# adding regions (Northeast, Midwest, South, and West)
merged_region = merge(merged_df_long,region[-1], by='state')

# region mean
only_region = merged_region[,-c(3,4)]
house_price_region=  only_region |> group_by(Region, Year) |> summarise(HousePrice = mean(HousePrice)) 

B = ggplot() + geom_line(data = house_price_region, aes(x = Year, y = HousePrice, group = Region, color = Region), alpha = 1)+ ggtitle("House price for each region over the year") + scale_x_continuous(breaks = c(1975, 1985, 1995, 2005, 2015, 2025 )) 
B
```

## states within a region
```{r}
West <- filter(merged_region, Region == "West")
South <- filter(merged_region, Region == "South")
Northeast <- filter(merged_region, Region == "Northeast")
Midwest <- filter(merged_region, Region ==  "Midwest")

W = ggplot(West, aes(y = HousePrice, x = Year, color = state)) +
  geom_line() + ggtitle("House price for West region over the year") + 
  scale_x_continuous(breaks = c(1975, 1985, 1995, 2005, 2015, 2025 )) 

S = ggplot(South, aes(y = HousePrice, x = Year, color = state)) +
  geom_line() + ggtitle("House price for South region over the year") + 
  scale_x_continuous(breaks = c(1975, 1985, 1995, 2005, 2015, 2025 ))

NE = ggplot(Northeast, aes(y = HousePrice, x = Year, color = state)) +
  geom_line() + ggtitle("House price for Northeast region over the year") + 
  scale_x_continuous(breaks = c(1975, 1985, 1995, 2005, 2015, 2025 ))

MW = ggplot(Midwest, aes(y = HousePrice, x = Year, color = state)) +
  geom_line() + ggtitle("House price for Midwest region over the year") + 
  scale_x_continuous(breaks = c(1975, 1985, 1995, 2005, 2015, 2025 ))

W
S
NE
MW

```



```


In the mid oughts, all regions drop due to the financial crash
1983 house crisis.
early 2000 housing bubles (2003-2007)
Also, all regions begin to reverse recession trends due to the economic recovery in the mid 2010s




### how to combine two different files and data analysis all together?


2. Population density and changes in house prices: Does present-day population density explain changes in house prices by state from 1975 to 2019? Are there outliers to the relationship, and if so, is there a principled reason to drop them? What does the relationship look like after dropping or downweighting outliers? Does the relationship vary by region? If so, how?

```{r}

pop00 = read.delim("censuspop2000.txt", header = TRUE, sep = "")
pop10 = read.delim("censuspop2010.txt", header = TRUE, sep = "")
pop19 = read.delim("censuspop2019.txt", header = TRUE, sep = "") 
# pop19, until row 52 : population


pop = c(sum(pop00$value), sum(pop10$value)) #, sum(pop19$value))
density = c(pop00$value, pop10$value)/pop



```





3. Changes in population and changes in house prices: Is there a relationship between changes in population and changes in house prices? To answer this, look at changes in each state over three time periods: 1990 to 2000, 2000 to 2010, and 2010 to 2019. Analyze the three time periods separately. Has the relationship changed over the three time periods? Are there variations by region?






4. Conclusion: What does all of this tell you about the relationship between house prices and population? Is there a plausible cause-and-effect story you can tell that’s consistent with the data and with common sense?





